---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data
```{r set options, echo=TRUE}
opts_chunk$set(echo=TRUE, results="asis")
```

```{r read data} 
activity <- read.csv("activity.csv", colClass=(c("numeric", "Date", "numeric")), na.strings="NA")
```

## What is mean total number of steps taken per day?

Here is a histogram of the total number of steps per day.
```{r historgram}
library(dplyr)
act_day <- group_by(activity, date)
act_totals <- summarize(act_day, total_steps=sum(steps))
hist(act_totals$total_steps, main="Total Steps Per Day", xlab="Steps", col=c("red"))
```

Here is the mean total number of steps taken per day.
```{r mean, results='asis'}
mean(act_totals$total_steps, na.rm=TRUE)
```

Here is the median total number of steps taken per day.
```{r median}
median(act_totals$total_steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

```{r average activity pattern} 
act_int <- group_by(activity, interval)
act_avg <- summarize(act_int, mean_steps=mean(steps, na.rm=TRUE))
plot(act_avg$interval, act_avg$mean_steps, type="l", main="Average Daily Activity Pattern", xlab="5-Minute Interval", ylab="Mean Steps")
``` 

The 5-minute interval, averaged across days, that contains the highest number of steps is:
```{r max steps}
max <- filter(act_avg, mean_steps==max(act_avg$mean_steps))
max
```

## Imputing missing values
The number of missing values is:
```{r missing values}
sum(is.na(activity$steps))
```


This code replaces the missing values with the average for each interval.
```{r replace missing values}
library(data.table)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm=TRUE))
setDT(activity)
replaced <- activity[,steps := impute.mean(steps), by = interval]
```

Here is a histogram of the total number of steps per day with the missing values filled in.
```{r NAs replaced historgram}
rep_day <- group_by(replaced, date)
rep_totals <- summarize(rep_day, total_steps=sum(steps))
hist(rep_totals$total_steps, main="Total Steps Per Day (NAs replaced)", xlab="Steps", col=c("darkblue"))
```

Here is the mean total number of steps taken per day with the missing values filled in.
```{r replaced meean}
mean(rep_totals$total_steps)
```

Here is the median total number of steps taken per day with the missing values filled in.
```{r replaced median}
median(rep_totals$total_steps)
```


Yes, replacing the missing values does have an effect on the data. The columns in the replaced histogram are a little taller because now they have more data to count and the mean and median are the same in the new histogram.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays versus weekends}
Weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity$day <- factor((weekdays(activity$date) %in% Weekdays), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

wkday <- filter(activity, day=="weekday")
wkend <- filter(activity, day=="weekend")
intwkday <- group_by(wkday, interval)
intwkend <- group_by(wkend, interval)
wkdayavg <- summarize(intwkday, mean_steps=mean(steps, na.rm=TRUE))
wkendavg <- summarize(intwkend, mean_steps=mean(steps, na.rm=TRUE))
par(mfrow=c(2,1))
plot(wkdayavg$interval, wkdayavg$mean_steps, type="l", main="Average Weekday Activity Pattern", xlab="Interval", ylab="Mean Steps")
plot(wkendavg$interval, wkendavg$mean_steps, type="l", main="Average Weekend Activity Pattern", xlab="Interval", ylab="Mean Steps")
```
